---
  title: "Homework_4"
author: "Mary Lewis"
date: "October 16, 2016"
output: pdf_document
---



```{r warning=FALSE, message=FALSE}

rm(list=ls())
library(car)
library(lmtest)
library(sandwich)
library(data.table)

load("C:/athletics.RData")
```
Question 1)
Conduct a quick examination and EDA of the dataset.
Examine the variables of interest
```{r}
ath1 <- data.table(data)
str(ath1)
desc
nrow(ath1)
summary(ath1)
hist(ath1$apps)
```
Log improves the apps distribution
Many schools have been in a bowl, but very few have a btitle or been in the final four

```{r }
hist(log(ath1$apps))
hist(ath1$bowl)
hist(ath1$btitle)
hist(ath1$finfour)
```

Question 2)
Which schools had the greatest increase and the greatest decrease in number of log applications?

Distribution of clapps looks almost normal, slight right skew, median is less than mean
Arizona has largest increase in clapps
Arkansas has largest decrease in clapps

```{r}

ath = merge(x = ath1[ath1$year == 1992, ], y = ath1[ath1$year == 1993, ], by = "school", 
            all=TRUE, suffixes = c("_92", "_93"))

ath$clapps<-ath$lapps_93 - ath$lapps_92
median(ath$clapps)
mean(ath$clapps)
hist(ath$clapps)
most<-ath$school[which.max(ath$clapps)]
most
min<-ath$school[which.min(ath$clapps)]
min

```


Question 3)
Similarly to above, create three variables, cbowl, cbtitle, and cfinfour, where each of these
variables represents the changes in the three athletic success variables.
```{r warning=FALSE}

ath$cbowl<-ath$bowl_93- ath$bowl_92
ath$ctitle<-ath$btitle_93 - ath$btitle_92
ath$cfinfour<-ath$finfour_93 - ath$finfour_92
```

Question 4)
We are interested in a population model,
lapps(i) = B0 + B1 cbowl(i) + B2 cbtitle(i) + B3 cfinfour(i) + a(i) + u(it)

The univ with the largest increase, arizona, has leverage at 0.5
Remove this and re-examine model
Result is that cBowl, which had been significant at 5% is no longer significant at 5% 
(Now it's probability density is about 10%)
==> Conclusion is to leave the row for Arizona in the dataset
 
```{r}
model1<-lm(ath$clapps ~ ath$cbowl + ath$ctitle + ath$cfinfour)
summary(model1)
plot(model1, 1)
plot(model1, 2)
plot(model1, 3)
plot(model1, 5)

ath2<-ath[-c(2), ]

model1a<-lm(ath2$clapps ~ ath2$cbowl + ath2$ctitle + ath2$cfinfour)
summary(model1a)
plot(model1a, 1)
plot(model1a, 2)
plot(model1a, 3)
plot(model1a, 5)
```


#4a) What additional assumption is needed for this population model to be causal? Write this in mathematical #notation and also explain it intuitively in English.

#For the model to be causal, the change in the error term as x changes must be zero.
Mathmatical expression: du/dx = 0
This means that as variables change for each school, changes in the error 
term should not be correlated with changes in those variables. In this case, 
the variables in the equation represent changes in athletic performance from 1991-1992.

4b) What additional assumption is needed for OLS to consistently estimate the first-difference model?
Write this in mathematical notation and also explain it intuitively in English. Comment on whether
this assumption is plausible in this setting.

The result of difference in differences, after taking the difference of 2 models at t1 and t2:

d/y(i) = B0 + B1 dx(i) + du(i) 

In order for OLS to consistently estimate the first difference model, we 
must assume that changes in the idiosyncratic error of an individual, i, over time, 
are independent of changes in x.

Mathmatical expression: du(i) / dx(i) = 0

The first-difference model itself has already accounted for variables 
that are consistent over time but are not included in the model (e.g. ability).
These are respresented as a(i) and cancel out measuring from t1 to t2.

However, changes in the error term for each individual between t1 and t2 
must be independent of any treatment effect (change in x) between t1 and t2. 

This is a plausible assumption for this model, assuming there was no sweeping policy 
change in application policy during the 2 years in question across all schools. 
For example, if a national policy made all applications free in 1993 (but not in 1992)
applications might have increased unexpectedly at certain schools not others in 1993, 
and this impact would occur in the error term at t2 but not t1. 


5) Null hypothesis: b1=b2=b3=0. The Wald test shows the model is not significant, 
so we can not reject the null hypothesis that indicator coefficients are all zero.
```{r}

waldtest(model1, vcov = vcovHC)

```
