---
title: "Section3_HW5_LewisM"
author: "Mary Lewis"
date: "October 23, 2016"
output: pdf_document
---

```{r}
rm(list=ls())
library(car)
library(lmtest)
library(sandwich)
library(data.table)

load("C:/voucher.RData")
prelim <- data.table(data)
str(prelim)
desc
nrow(prelim)
summary(prelim)
sum(prelim$choiceyrs>prelim$selectyrs)
```
Remove 2 rows where choice years are more than select years as this is not possible.
```{r}
voucher<-subset(prelim[prelim$choiceyrs<=prelim$selectyrs])
nrow(voucher)
```
Selectyrs normally distributed as would expect from random assignment
```{r}
hist(voucher$selectyrs)
```
Choiceyrs right skew
```{r}
hist(voucher$choiceyrs)
#summary(voucher$choiceyrs)
#summary(voucher$selectyrs)
```
Question 1: Of the 990 students in the sample, how many were never awarded a voucher? 
Answer : 467
How many had a voucher available for four years? 
Answer: 108
How many students actually attended a choice school for four years?
Answer: 56
```{r}
sum(voucher$select==0)
sum(voucher$selectyrs==4)
sum(voucher$choiceyrs==4)
```

Question 2: Run a simple regression of choiceyrs on selectyrs.
Are these variables related in the direction you expected? How strong is the relationship?

Directionally they are in line with what you would expect but only at .77. Ideally, with no attrition,
the coef would be 1. The rate of attrition is interpreted to be about 23% (100% - 77%)
The relatoinship is highly significant.
Is selectyrs a sensible IV candidate for choiceyrs? As mentioned in the description, it's not a completely
random assignment becuase a certain subset might be more likely to apply for a voucher.
It's exogenous to a greater extent than choice years but probably not truly exogenous.
```{r}
model1<-lm(voucher$choiceyrs ~ voucher$selectyrs)
summary(model1)
```
Question 3: Run a simple regression of mnce on choiceyrs. What do you find? Is this what you expected?
What happens if you add the variables black, hispanic, and female?
 
Answer: It is somewhat surprising that for each additional year at choice school, mnce score decreases by 1.8 points
```{r}
model2<-lm(voucher$mnce ~ voucher$choiceyrs)
summary(model2)
```
Adding all 3 (last model below), choice years is no longer significant, whereas
being black or hispanic is practically and statistically significant and accounts 
for the negative relationship.
Being female has a positive relationship to mnce but is not statistically significant. 
```{r}
model3<-lm(voucher$mnce ~  voucher$choiceyrs + voucher$black)
summary(model3)
model4<-lm(voucher$mnce ~ voucher$choiceyrs + voucher$hispanic)
summary(model4)
model5<-lm(voucher$mnce ~ voucher$choiceyrs + voucher$female)
summary(model5)

model_all<-lm(voucher$mnce ~  voucher$choiceyrs + voucher$black + voucher$hispanic + voucher$female)
summary(model_all)
```
4) The equation omits important variables that are likely to impact
the number of choiceyears, such as family income, parents level of education, etc.
Thus the error term is not uncorrelated with choiceyears.

5)
Using select years as the IV, mnce still is negative, again not significant.
for each increase in  select year, mnce drops, but to a lesser extent than choice years.
Practically and statistically, the coefs on the other variables are essentially the same as they were
using choice years, but the coef on IV (select years) is less than choice years. (causes less of a decline)

This seems to confirm that choice years is endogenous.
```{r}
model_IV<-lm(voucher$mnce ~  voucher$selectyrs + voucher$black + voucher$hispanic + voucher$female)
summary(model_IV)
```
6) Estimate equation #4 , include mnce90 variable. Test with OLS and IV. Compare the results for B1.

When adding prior performance, mnce90, Choice years is not significant, coef is 0.4.
Mnce90 is highlty precdictive of mnce at 0% significance (.62). Black is practically & significantly
significant.
```{r}
model_all_90 <-lm(voucher$mnce ~  voucher$choiceyrs + voucher$mnce90 + voucher$black 
                  + voucher$hispanic + voucher$female)
summary(model_all_90)
```
Switching to IV, Select years, mnce is impacted 1.3 pts, per year and is significant at 5%.
Over a four year period, this equates to 5 points. We would need to
know more about mnce scale and scoring
to know if 5 points is practically significant.
Mnce90 and black have coefs in the same range as when not using the IV.
```{r}
model_all_90_IV <-lm(voucher$mnce ~  voucher$selectyrs + voucher$mnce90 + voucher$black 
                     + voucher$hispanic + voucher$female)
summary(model_all_90_IV)
```
7) Estimate choiceyrs and selectyrs using categorical variables.
None of the coefs on the categorical vars are significant.
Students with selectyrs = 3 are the only students with negative coef
relation to mnce relative to those who received vouchers (other selectyrs) as wellas those that did not
receive voucher (selectyrs=0).
Findings suggest that not attending first year at a choice school has a negative impact,
including that it might increase the attrition rate for that same group..
```{r}
model_cat_choice<-lm(voucher$mnce ~  choiceyrs1 + choiceyrs2 + choiceyrs3 + choiceyrs4 
                     + black + hispanic + female, data=voucher)
summary(model_cat_choice)

model_cat_select<-lm(voucher$mnce ~   selectyrs1 + selectyrs2 + selectyrs3 + selectyrs4 
                     + black + hispanic + female, data=voucher)
summary(model_cat_select)
```
